### 2016-05-21 Paper_v2.tex
Updated version

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Reduced order modeling and analysis of the human complement subsystem}}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Wei Dai$^{\#}$, Mason Minot$^{\#}$, and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Reduced order model of complement}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{PLoS~ONE}}
\vspace{0.5in}
{\par \centering $^{\#}$ Denotes equal contribution}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Complement is a central part of innate immunity and plays a very significant role in regulating the inflammatory response.
In this study, we present an effective biochemical network modeling approach in building a reduced order model to study a complex biochemical network, the human complement system. The key innovation of our approach is the use of simple equations to capture the behavior of a complex biochemical network.  We developed a hybrid modeling approach which combines ODEs and logical rules to model biochemical processes for which a complete mechanistic understanding is missing. Using this modeling framework we  incorporated computational simulation and experimental validation of the lectin and alternative pathway to examine C3 and C5 convertase formation and activity. The reduced order model consisted of only 18 differential equations with 28 kinetic and control parameters. Thus, the model was several orders of magnitude smaller and includes more pathways than comparable purely ODE models in the literature. We estimated the model parameters from in vitro human complement time-course experiments of C3a and C5a from Morad and coworkers \cite{morad2015time}, in the presence and absence of zymosan, and without the classical pathway. We then compared the model predictions with C3a and C5a data sets, for alternative and lectin pathway dynamics, that were not used for model training. Once validated, we performed sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. The reduced order hybrid approach produced a surprisingly predictive human complement model, much similar to the study on human coagulation using the same modeling framework \cite{sagar2015dynamic}. Taken together, the combined analysis of alternate and lectin pathways along with the incorporation of the downstream reactions involving C5 convertase elucidated new insight into roles of governing parameters that mediate the complement system. Deeply understanding how the main governing parameters change with respective to the different combinations of pathway activation will greatly aid in development of drugs for strategic therapeutic targets. Due to the low computational cost relative to the existing models and accuracy of our predictions, we believe that our reduced order complement network is the first step towards building a computation toolbox for treatment analysis or clinical screening in the medical field.

%We further tested the predictive power of the coagulation model parameters against data not used in training, and found good agreement between simulations and experimental measurements. Lastly, we tested the performance of DOPS on commonly used test functions for global optimization and on published biochemical parameter estimation benchmark problems. For the wide range of problems that we considered, DOPS outperformed other meta-heuristic approaches despite a limited number of function evaluations.

@@

% Extra abstract
% Mathematical modeling of biological systems with multiple feed back loops is one such area where parameter estimation is a difficult non-linear optimization problem. This difficulty is further compounded when dealing with parameter vectors of high dimensions.

%In this study, we present the dynamically dimensioned particle
%a novel meta-heuristic approach that combines a variant of particle swarm optimization (PSO) along with dynamically dimensioned search (DDS) to obtain near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations.
%We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions. We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone. The new algorithm outperforms all the other meta-heuristics on the coagulation model. The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data. We also performed this comparison on commonly used test functions (Ackley and Rosenbrock) for global optimization and found the same behavior. Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach. We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.

@@

The activation of complement and formation of C5a and MAC complex is regulated at different points through a number of plasma and host cell proteins.
The initiation of the classical pathway through the attachment of C1 to an antibody is controlled by the C1 Inhibitor (C1-Inh), a protease inhibitor belonging to the serpin superfamily. C1-Inh irreversibly binds to and deactivates the active subunits of component C1 to prevent spontaneous fluid phase and chronic activation of complement \cite{walker1995complement}. The serum and host-tissue regulation of the upstream elements of the complement system is also achieved through the binding of C4 binding protein (C4BP) to C4b and through the binding of factor H to C3b \cite{blom2001structural}. These proteins are also capable of binding their respective components in the convertase form. Membrane cofactor protein (MCP or CD46) possesses a cofactor activity for C4b and C3b, which protects the host from self-activation of complement \cite{riley2004cd46}.  Delay accelerating factor (DAF or CD55) is able to recognize and dissociate both convertases \cite{lukacik2004complement}. MAC is inhibited by vitronectin and clusterin in the plasma and CD59 at the host surface.
Proteins C3a, C4a, and C5a are inactivated or reduced in activity by carboxypeptidase-N \cite{liszewski1995control}.

Given the complexity and importance, developing mathematical models of complement are crucial to understanding its dynamics. Complement models have typically been formulated as linear or non-linear Ordinary Differential Equation (ODE) systems. Hirayama  et al. (ref) used a system of linear ODEs to model the classical pathway of complement. Korotaevskiy and co-workers (ref) built a theoretical model of complement using a system of non-linear ODEs that included classical, lectin and alternate pathways. However both these studies involve no validation studies with experimental data. Liu et al used analyzed the formation of classical and lectin C3 convertases and the regulatory role of C4BP using a system of 45 non-linear ODEs with 85 parameters.  Recently, Zewde and co-workers built a detailed mechanistic model of alternative complement activation was built using 107 ODEs and 74 kinetic parameters (Ref). This model delineated the response of complement on a host cell and a foreign antigen. However, these previous models were largely based upon mechanistic knowledge.  However given the complexity of complement and its interactions with other networks like coagulation, autonomous nervous system, adaptive immunity it is unfeasible and computationally expensive to build such large mechanistic models. In addition is much more difficult to experimentally interrogate the response of various complement proteins under different conditions. This also presents with the problem of estimation of a large number of parameters with little or no experimental data. Thus there exists a need to reduce the mechanistic complexity while capturing dynamics of complement accurately.

@@

>>>>

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Reduced order modeling and analysis of the human complement subsystem}}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Wei Dai$^{\#}$, Mason Minot$^{\#}$, and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Reduced order model of complement}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{PLoS~ONE}}
\vspace{0.5in}
{\par \centering $^{\#}$ Denotes equal contribution}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Complement is a central part of innate immunity and plays a significant role in regulating the inflammatory response. In this study, we build a reduced order model of complement to study the human complement system.
The key novelty of our approach is the use of Ordinary Differential Equations (ODEs) along with logical rules to capture the behavior of a complex biochemical network.  Using this framework we constructed a model of complement that analyzed the dynamics of C3a and C5a when initiated through the alternate and lectin pathways. The reduced order model consisted of only 18 differential equations with 28 kinetic and control parameters. Thus, the model was an order of magnitude smaller that any existing model of complement that includes alternate and lectin pathways.

%and includes more pathways than comparable purely ODE models in the literature. We estimated the model parameters from in vitro human complement time-course experiments of C3a and C5a from Morad and coworkers, in the presence and absence of zymosan, and without the classical pathway.

%We then compared the model predictions with C3a and C5a data sets, for alternative and lectin pathway dynamics, that were not used for model training. Once validated, we performed sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. The reduced order hybrid approach produced a surprisingly predictive human complement model, much similar to the study on human coagulation using the same modeling framework \cite{sagar2015dynamic}. Taken together, the combined analysis of alternate and lectin pathways along with the incorporation of the downstream reactions involving C5 convertase elucidated new insight into roles of governing parameters that mediate the complement system. Deeply understanding how the main governing parameters change with respective to the different combinations of pathway activation will greatly aid in development of drugs for strategic therapeutic targets. Due to the low computational cost relative to the existing models and accuracy of our predictions, we believe that our reduced order complement network is the first step towards building a computation toolbox for treatment analysis or clinical screening in the medical field.

@@

@@

The activation of complement and formation of C5a and MAC complex is regulated at different points through a number of plasma and host cell proteins.
The initiation of the classical pathway through the attachment of C1 to an antibody is controlled by the C1 Inhibitor (C1-Inh), a protease inhibitor belonging to the serpin superfamily. C1-Inh irreversibly binds to and deactivates the active subunits of component C1 to prevent spontaneous fluid phase and chronic activation of complement \cite{walker1995complement}. The serum and host-tissue regulation of the upstream elements of the complement system is also achieved through the binding of C4 binding protein (C4BP) to C4b and through the binding of factor H to C3b \cite{blom2001structural}. These proteins are also capable of binding their respective components in the convertase form. Membrane cofactor protein (MCP or CD46) possesses a cofactor activity for C4b and C3b, which protects the host from self-activation of complement \cite{riley2004cd46}.  Decay accelerating factor (DAF or CD55) is able to recognize and dissociate both convertases \cite{lukacik2004complement}. MAC is inhibited by vitronectin and clusterin in the plasma and CD59 at the host surface.
Proteins C3a, C4a, and C5a are inactivated or reduced in activity by carboxypeptidase-N \cite{liszewski1995control}.

Given the complexity and importance, developing mathematical models of complement are crucial to understanding its dynamics. Complement models have typically been formulated as linear or non-linear Ordinary Differential Equation (ODE) systems. Hirayama  et al. (ref) used a system of linear ODEs to model the classical pathway of complement. Korotaevskiy and co-workers (ref) built a theoretical model of complement using a system of non-linear ODEs that included classical, lectin and alternate pathways. However both these studies involve no validation studies with experimental data. Liu et al analyzed the formation of classical and lectin C3 convertases and the regulatory role of C4BP using a system of 45 non-linear ODEs with 85 parameters.  Recently, Zewde and co-workers built a detailed mechanistic model of alternative complement activation was built using 107 ODEs and 74 kinetic parameters (Ref). This model delineated the response of complement on a host cell and a foreign antigen. However, these previous models were largely based upon mechanistic knowledge.  However given the complexity of complement and its interactions with other networks like coagulation, autonomous nervous system, adaptive immunity it is unfeasible and computationally expensive to build such large mechanistic models. In addition is much more difficult to experimentally interrogate the response of various complement proteins under different conditions. This also presents with the problem of estimation of a large number of parameters with little or no experimental data. Thus there exists a need to reduce the mechanistic complexity while capturing dynamics of complement accurately.

@@

In this study, we present an effective biochemical network modeling approach in building a reduced order model to study a complex biochemical network, the human complement system. The key innovation of our approach is the use of simple equations to capture the behavior of a complex biochemical network.  We developed a hybrid modeling approach which combines ODEs and logical rules to model biochemical processes for which a complete mechanistic understanding is missing or is complex. Using this modeling framework we  incorporated computational simulation and experimental validation of the lectin and alternative pathway to examine C3 and C5 convertase formation and activity. The reduced order model consisted of only 18 differential equations with 28 kinetic and control parameters. Thus, the model was several orders of magnitude smaller and includes more pathways than comparable purely ODE models in the literature. We estimated the model parameters from in vitro human complement time-course experiments of C3a and C5a from Morad and coworkers \cite{morad2015time}, in the presence and absence of zymosan, and without the classical pathway. We then compared the model predictions with C3a and C5a data sets, for alternative and lectin pathway dynamics, that were not used for model training. Once validated, we performed sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. The reduced order hybrid approach produced a surprisingly predictive human complement model, much similar to the study on human coagulation using the same modeling framework \cite{sagar2015dynamic}. Taken together, the combined analysis of alternate and lectin pathways along with the incorporation of the downstream reactions involving C5 convertase elucidated new insight into roles of governing parameters that mediate the complement system. Deeply understanding how the main governing parameters change with respective to the different combinations of pathway activation will greatly aid in development of drugs for strategic therapeutic targets. Due to the low computational cost relative to the existing models and accuracy of our predictions, we believe that our reduced order complement network is the first step towards building a computation toolbox for treatment analysis or clinical screening in the medical field.

%%%The start of the second paragraph

This is the first ever model of complement that combines the alternate and non-alternate pathways and validates the model on experimental data.
The formation of C3a without any initiator is controlled by the tick-over mechanism.

Though it has been known that complement is a key player in regulating immune response and inflammation, very few mathematical models have been built to understand the dynamics of this network. To our knowledge this is the first model of complement that combines different pathways of initiation and validates the dynamics of C3a and C5a on experimental data. The cleavage to C4 and C2 to form the classical/lectin C3 convertase involves a number of complex steps. The formation of this convertase and thereby the formation of C3a and C3b was modeled with  45 non-linear ODEs with 85 parameters by Liu and co-workers. The hybrid modeling framework allowed us to model the lectin mediated formation of C3 convertase using only 3 ODEs. We were able to successfully capture the formation of C3a at the maximum initiator (zymosan) concentration and also were able to qualitatively predict the dynamics of lectin mediated C3a formation at different initiator levels. The model is also surprisingly accurate in capturing the dynamics of C3a and C5a formed from the alternate pathway.
